---
title: "Plotting in ratlas"
output:
  bookdown::html_document2:
    number_sections: false
    toc: true
pkgdown:
  as_is: true
resource_files:
  - images/
vignette: >
  %\VignetteIndexEntry{Plotting in ratlas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, child="children/chunk-options.txt"}
```

```{r setup, include = FALSE}
library(ratlas)
library(dplyr)
library(ggplot2)

library(scales)
library(viridis)
library(dichromat)
library(colorspace)
```

This vignette walks through how to create consistent graphics that meet ATLAS formatting standards using [{ggplot2}](https://ggplot2.tidyverse.org). The {ggplot2} package uses the grammar of graphics to create elegant and reproducible plots that are publication ready. The {ratlas} package provides an extension to {ggplot2} by customizing the default theme to match brand requirements.

## Introduction to ggplot2

A full tutorial of how to use {ggplot2} is beyond the scope of this vignette. If you are new to {ggplot2}, we recommend you start with these resources:

* [Data Visualisation](https://r4ds.had.co.nz/data-visualisation.html) in *R for Data Science*, by Hadley Wickham & Garrett Grolemund
* [*Data Visualization*](https://socviz.co/), by Kieran Healy
* [*R Graphics Cookbook*](https://r-graphics.org), by Winston Chang

These resources walk through how {ggplot2} works, and provide examples with code for creating different types of plots. For a more in depth description of how {ggplot2} works, we recommend [*ggplot2: Elegant Graphics for Data Analysis*](https://ggplot2-book.org), by Hadley Wickham. The second edition can be purchased from [Amazon](https://www.amazon.com/ggplot2-Elegant-Graphics-Data-Analysis/dp/331924275X/ref=sr_1_1?keywords=ggplot2&qid=1568726727&sr=8-1); however, the third edition is currently under development, and can be accessed for free at [ggplot2-book.org](https://ggplot2-book.org).

To demonstrate how to use the ATLAS theme for {ggplot2}, we will use the following example plot of the `mtcars` data set (Figure \@ref(fig:init-plot)).

(ref:init-plot-cap) Default {ggplot2} theme.

```{r init-plot, fig.cap = "(ref:init-plot-cap)"}
library(ggplot2)

ggplot(mtcars, aes(x = mpg, y = disp)) +
  geom_point(aes(color = factor(cyl)))
```


## Using `theme_atlas()`

The ATLAS {ggplot2} theme can be applied by adding a `ratlas::theme_atlas()` layer to plot, just as you would add any other theme. The result is shown in Figure \@ref(fig:theme-atlas-ex).

(ref:theme-atlas-ex-cap) The result of applying `ratlas::theme_atlas()` to a plot.

```{r theme-atlas-ex, fig.cap = "(ref:theme-atlas-ex-cap)"}
library(ratlas)

ggplot(mtcars, aes(x = mpg, y = disp)) +
  geom_point(aes(color = factor(cyl))) +
  theme_atlas()
```

There are a few points to draw attention to when comparing Figure \@ref(fig:theme-atlas-ex) back to Figure \@ref(fig:init-plot). First, the gray  background has been removed, and the grid lines have been changed from white to gray. Second, the legend has been moved from a vertical orientation on the right-hand side of the plot to a horizontal orientation below the plot. Finally, the font has been changed to Arial Narrow, which should be available on all systems and provides a more modern look.

## Changing the font

Although Arial Narrow provides a modern font available on all systems, it is often desirable to have a font that matches the brand requirements. The official font of the University of Kansas and ATLAS is [Gotham](https://brand.ku.edu/guidelines/design/typography). However, this font is expensive, and thus inaccessible to most users. [Montserrat](https://fonts.google.com/specimen/Montserrat) is a very similar font that is available for free through Google Fonts. We recommend that you load this font with the {showtext} package. This will load fonts that are available online through Google Fonts, and then tell R to render text using {showtext} by calling the `showtext_auto()` function.

```{r showtxtfnt, eval = FALSE}
library(showtext)
## Loading Google fonts (https://fonts.google.com/)
font_add_google("Montserrat", "montserrat")

## Automatically use showtext to render text
showtext_auto()
```

Once the Montserrat font has been loaded, you can apply it to the ATLAS theme using `ratlas::theme_atlas(base_family = "Montserrat")`.

```{r theme-atlas-ms-ex-build, eval = FALSE}
ggplot(mtcars, aes(x = mpg, y = disp)) +
  geom_point(aes(color = factor(cyl))) +
  theme_atlas(base_family = "Montserrat")
```

## Colors

One potential issue with graphics created with {ggplot2} is that the default color scales are not colorblind friendly. For example, Figure \@ref(fig:disc-cvd) shows our working example simulated with different types of color blindness.

(ref:disc-cvd-cap) The default {ggplot2} color palette for discrete scales with different types of color blindness.

```{r colorblindr-funcs, include = FALSE}
# This is a placeholder until colorblindr is on CRAN. When that happens,
# colorblindr can be added to suggests, and we can just use
# colorblindr::cvd_grid()
edit_colors <- function(plot = last_plot(), colfun = passthrough, fillfun = NULL, ...)
{
  # convert to grob if necessary
  if (!methods::is(plot, "grob")) {
    plot <- cowplot::plot_to_gtable(plot)
  }

  if (is.null(fillfun)) {
    fillfun = colfun
  }
  edit_grob_colors(plot, colfun, fillfun, ...)
}

edit_grob_colors <- function(grob, colfun, fillfun, ...)
{
  if (!is.null(grob$gp)) {
    if (!is.null(grob$gp$col)) {
      grob$gp$col <- colfun(grob$gp$col, ...)
    }
    if (!is.null(grob$gp$fill)) {
      grob$gp$fill <- fillfun(grob$gp$fill, ...)
    }
  }

  if (!is.null(grob$grobs)) {
    grob$grobs <- lapply(grob$grobs, edit_grob_colors, colfun, fillfun, ...)
  }

  if (!is.null(grob$children)) {
    grob$children <- lapply(grob$children, edit_grob_colors, colfun, fillfun, ...)
  }

  if (methods::is(grob, "rastergrob")) {
    grob <- edit_rastergrob_colors(grob, colfun, ...)
  }

  grob
}

edit_rastergrob_colors <- function(grob, colfun, ...)
{
  rasternew <- colfun(c(grob$raster), ...)
  dim(rasternew) <- dim(grob$raster)
  class(rasternew) <- class(grob$raster)
  grid::editGrob(grob, raster = rasternew)
}

cvd_grid <- function(plot = last_plot(), severity = 1)
{
  deut <- function(c) colorspace::deutan(c, severity)
  p1 <- edit_colors(plot, deut)

  prot <- function(c) colorspace::protan(c, severity)
  p2 <- edit_colors(plot, prot)

  trit <- function(c) colorspace::tritan(c, severity)
  p3 <- edit_colors(plot, trit)

  des <- function(c) colorspace::desaturate(c, severity)
  p4 <- edit_colors(plot, des)

  cowplot::plot_grid(p1, p2, p3, p4, scale = 0.9, hjust = 0, vjust = 1,
                     labels = c("Deutanomaly", "Protanomaly", "Tritanomaly", "Desaturated"),
                     label_x = 0.01, label_y = 0.99, label_size = 12, label_fontface = "bold")
}
```

```{r disc-cvd, echo = FALSE, fig.asp = 1.618, fig.cap = "(ref:disc-cvd-cap)"}
p <- ggplot(mtcars, aes(x = mpg, y = disp)) +
  geom_point(aes(color = factor(cyl))) +
  theme_atlas()

cvd_grid(p)
```

Figure \@ref(fig:disc-cvd) shows that most types of color blindness make it hard to distinguish between the different level of the `cyl` variable when using the default color palette. The palette for continuous variables has similar problems. Figure \@ref(fig:cont-ex) shows the default color palette for continuous variables in {ggplot2}.

(ref:cont-ex-cap) Default color palette for continuous variables in {ggplot2}.

```{r for_repeat, include = FALSE}
n_col <- 128
img <- function(obj, nam) {
  image(1:length(obj), 1, as.matrix(1:length(obj)), col = obj, main = nam,
        ylab = "", xaxt = "n", yaxt = "n",  bty = "n")
}
```

```{r cont-ex, echo = FALSE, fig.asp = 0.15, fig.cap = "(ref:cont-ex-cap)"}
par(mfrow=c(1, 1), mar=rep(1, 4))
img(rev(seq_gradient_pal(low = "#132B43", high = "#56B1F7",
                         space = "Lab")(seq(0, 1, length=n_col))), "")
```

The simulated color blindness for this scale is shown in Figure \@ref(fig:cont-cvd). Here, several types of color blindness result in large sections of the palette where it is difficult to differentiate values.

(ref:cont-cvd-cap) The default {ggplot2} color palette for continuous scales with different types of color blindness.

```{r cont-cvd, echo = FALSE, fig.asp = 0.5, fig.cap = "(ref:cont-cvd-cap)"}
par(mfrow=c(4, 1), mar=rep(1, 4))
img(dichromat(rev(seq_gradient_pal(low = "#132B43", high = "#56B1F7",
                                   space = "Lab")(seq(0, 1, length=n_col))),
              "deutan"), "Deutanomaly")
img(dichromat(rev(seq_gradient_pal(low = "#132B43", high = "#56B1F7",
                                   space = "Lab")(seq(0, 1, length=n_col))),
              "protan"), "Protanomaly")
img(dichromat(rev(seq_gradient_pal(low = "#132B43", high = "#56B1F7",
                                   space = "Lab")(seq(0, 1, length=n_col))),
              "tritan"), "Tritanomaly")
img(desaturate(rev(seq_gradient_pal(low = "#132B43", high = "#56B1F7",
                                    space = "Lab")(seq(0, 1, length=n_col)))),
    "Desaturated")
```

### Using color blind safe palettes

For discrete color scales, the {ratlas} package includes the Okabe Ito color palette developed by [Masataka Okabe and Kei Ito](https://jfly.uni-koeln.de/color/). This can be used with the `ratlas::scale_color_okabeito()` function.

```{r okabeito-compare, out.width = "47%", fig.asp = 1.618, fig.show = "hold", fig.cap = "Comparison plot using the default (left) and Okabe Ito (right) discrete color palettes."}
ggplot(mtcars, aes(x = mpg, y = disp)) +
  geom_point(aes(color = factor(cyl)), size = 3) +
  theme_atlas()

ggplot(mtcars, aes(x = mpg, y = disp)) +
  geom_point(aes(color = factor(cyl)), size = 3) +
  scale_color_okabeito() +
  theme_atlas()
```

For continuous color scales, the viridis family of color palettes should be used. These palettes were created by [Stéfan van der Walt](https://github.com/stefanv) and [Nathaniel Smith](https://github.com/njsmith) and are shown in Figure \@ref(fig:show-viridis).

```{r show-viridis, echo = FALSE, fig.asp = 0.6, fig.cap = "The viridis color scales."}
par(mfrow=c(5, 1), mar=rep(1, 4))
img(rev(viridis(n_col)), "Viridis")
img(rev(magma(n_col)), "Magma")
img(rev(plasma(n_col)), "Plasma")
img(rev(inferno(n_col)), "Inferno")
img(rev(cividis(n_col)), "Cividis")
```

These color scales are perceptually uniform, perform well for a variety of colorblindness, and maintain the ability to discriminate across the perceptual spectrum when rendered or printed in gray scale. These color scales can be applied using `ggplot2::scale_color_viridis_c()`. Figure \@ref(fig:viridis-compare) shows and example of using the the viridis color palette. There is also a discrete option for the viridis color palette (`ggplot2::scale_color_viridis_d()`); however, the Okabe Ito palette is preferred for discrete scales.

```{r viridis-compare, out.width = "47%", fig.asp = 1.618, fig.show = "hold", fig.cap = "Comparison plot with default (left) and viridis (right) continuous color palettes."}
ggplot(faithfuld, aes(x = eruptions, y = waiting)) +
  geom_raster(aes(fill = density)) +
  theme_atlas()

ggplot(faithfuld, aes(x = eruptions, y = waiting)) +
  geom_raster(aes(fill = density)) +
  scale_fill_viridis_c() +
  theme_atlas()
```

### Brand colors

In additional to palettes that are color blind safe, the {ratlas} package also include brand specific color palettes. For example, a plot can be created using the official ATLAS color palette by using the `ratlas::scale_color_atlas()` function, as shown in Figure \@ref(fig:atlas-colors).

(ref:atlas-colors-cap) Plot with the official ATLAS colors using `ratlas::scale_color_atlas()`.

```{r atlas-colors, fig.cap = "(ref:atlas-colors-cap)"}
ggplot(mtcars, aes(x = mpg, y = disp)) +
  geom_point(aes(color = factor(cyl))) +
  scale_color_atlas() +
  theme_atlas()
```

## Setting a global theme

If you have several plots in a document, it can be a hassle to always remember to add `+ theme_atlas()` to every plot and to switch to a non-default color palette. To make the formatting of plots easier, {ratlas} includes a wrapper function, `ratlas::set_theme()` which will change the {ggplot2} defaults so that `theme_atlas()` is always used, and the default color scales are changed to the Okabe Ito palette for discrete scales and the viridis palette for continuous scales.

For example, once `set_theme()` has been called, the `scale_color_okabeito()` and `theme_atlas()` lines no longer need to be added. Instead, they are applied by default (Figure \@ref(fig:default-theme)). 

(ref:default-theme-cap) Plot generated after using `set_theme()` to set new defaults globally.

```{r default-theme, fig.cap = "(ref:default-theme-cap)"}
set_theme()

ggplot(mtcars, aes(x = mpg, y = disp)) +
  geom_point(aes(color = factor(cyl)))
```

The `set_theme()` function also provides flexibility for which defaults are set. For example, to use the Montserrat font by default, use `set_theme(font = "Montserrat")`. See `?set_theme()` for all options.

## Saving plots

Finally, it is often useful to save plots to a file. This can be done with `ratlas::ggsave2()`. This is a wrapper around `ggplot2::ggsave()`. The `ggsave2()` version applies some defaults to all ATLAS plots. Specifically, the plot labels and titles are automatically spell checked, the aspect ratio of plots is set to a reasonable default (rather than using the current size of your graphics device), and embeds the fonts when saving to a PDF format, allowing the plots to render correctly on any system.

```{r eval = FALSE}
p <- ggplot(mtcars, aes(x = mpg, y = disp)) +
  geom_point()

ggsave2(plot = p, filename = "my-plot.png", path = "where/to/save")
```

The `ggsave2()` function also flips the order of the `filename` and `plot` arguments from the `ggplot2::ggsave()` function. This allows calls to the function to be chained together, making it easier to save the sample plot in multiple formats using the `%>%` operator.

```{r eval = FALSE}
p %>%
  ggsave2(filename = "my-plot.png", path = "where/to/save") %>%
  ggsave2(filename = "my-plot.pdf", path = "where/to/save")
```

## Keep Learning

There are many resources available for learning more about how to make data visualizations with R and {ggplot2}. Here are a few **free** resources that I have found particularly helpful:

* For an introduction to {ggplot2}, see [Data Visualisation](https://r4ds.had.co.nz/data-visualisation.html), from *R for Data Science*, by Hadley Wickham and Garrett Grolemund
* For simple recipes for creating different types of plots with {ggplot2}, see the [*R Graphics Cookbook*](https://r-graphics.org/), by Winston Chang
* For more details on how to customize {ggplot2} graphics to fit your specific use case, see [*Data Visualization: A Practical Introduction*](https://socviz.co/), by Kieran Healy
* Finally, for a discussion of best principles for data visualization, see [*Fundamentals of Data Visualization*](https://clauswilke.com/dataviz/), by Claus Wilke
