---
title: "Plotting in ratlas"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{plotting-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.retina = 3,
  fig.width = 6,
  fig.asp = 0.618,
  fig.align = "center",
  out.width = "90%"
)

# Notes to self

# file structure has been made parallel to report-writing-vignette
  # make changes commensurate with being able to pass Github checks
# write Rmd vignette file
```

```{r setup, include = FALSE, eval = TRUE}
# see which of these I need and add to suggests in DESCRIPTION if needed
library(ratlas)
library(dplyr)
library(ggplot2)
library(english)
library(knitr)
library(fivethirtyeight)

options(warn = -1)
```

# Introduction to ggplot2

Figures are created using the ggplot2 package which is a part of the tidyverse 
family of packages. Figures are created using multiple steps and layers. The 
first step is to import the data into ggplot. This can be done using the 
following code:

```{r first_layer}
iris %>%
  ggplot()
```

The next step is to add layers to the graph. The first layer specifies the 
aesthetics. Within the ggplot function, the parameter *aes* corresponds to the 
aesthetics that are specified. For this graph, let's create a scatterplot with 
petal length on the x-axis, petal width on the y-axis, and the color of each 
point will correspond with species. The aesthetics of such a graph are specified
in the following code:

```{r second_layer}
iris %>%
  ggplot(aes(x = Petal.Length, y = Petal.Width, color = Species))
```

While the above code specifies how to map the graph aesthetics, no points are 
plotted on the coordinate plane. This is because the second layer involved in 
creating a figure is to add a geometric object layer that corresponds to the 
type of graph that is desired. In this case, we will use `geom_point` since we 
want to create a scatterplot. The geometric object layers are added using a *+* 
sign from the line above. The code below provides an example of adding the 
`geom_point` geometric object layer.

```{r third_layer, eval = TRUE}
iris %>%
  ggplot(aes(x = Petal.Length, y = Petal.Width, color = Species)) +
  geom_point()
```

In the graph above, you will notice that there are no parameters for the 
`geom_point` layer. This is done for two reasons. First, all of the global 
parameters specified in the `ggplot(aes(...))` layer are passed on to subsequent
layers. Thus, `geom_point` is run using the global parameters. Second, there 
were no necessary local changes, and hence no new parameters were necessary.

Geometric objects can also be layered on one another. The graph below adds a 
regression line using a `geom_smooth` object:

```{r fourth_layer, eval = TRUE}
iris %>%
  ggplot(aes(x = Petal.Length, y = Petal.Width, color = Species)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE)
```

Note, however, that adding the `geom_smooth` object actually provided three 
regression lines, one for each species of iris. This is because the global 
settings  are specified to color each species differently. In order to create a 
single regression line, the color aesthetic must be specified locally for the 
`geom_point` object. This will allow the points to be colored according to 
species, while drawing a single regression line through the points. Furthermore, 
the default color of the regression line will be overridden with 
`color = "black"` in the `geom_smooth` geometric object.

```{r iris-figure, eval = TRUE}
iris %>%
  ggplot(aes(x = Petal.Length, y = Petal.Width)) +
  geom_point(aes(color = Species)) + 
  geom_smooth(method = "lm", se = FALSE, color = "black")
```

## Facets

The example graph for the iris data provides an example of creating a single graph to display data. Now consider a second example which will graph the unemployement rate by median income of all college major categories using the `college_all_ages` dataset that comes with the `fivethirtyeight` package.

```{r college-ex, eval = TRUE}
college_all_ages %>% 
  ggplot(aes(x = unemployment_rate, y = median)) +
  geom_point(aes(color = major_category), show.legend = FALSE)
```

While the above graph contains a lot of information, it's rather convoluted because all of the college major categories are included in the same graph. Facets all for the data to be split based on one of the variables. 

There are two primary types of facets. First, there is `facet_grid()`, which splits the graph into a grid based on up to two chosen variables. One variable will specify how the graph is to be split into columns, and the other variable will specify how the graph will be split into rows. As will be shown in the example, one or two variables can be specified for defining the grid. The only consequence is the shape of the grid. Figure \@ref(fig:facet-grid-ex) shows the effect of using `facet_grid()` with the `major_category` variable. More help can be found by typing `?facet_grid` into the console. 

For simplicity, the following graph has been restricted to the Law & Public Policy, Communications & Journalism, Engineering, and Computers & Mathematics major categories.

```{r facet-grid-ex, eval = TRUE}
college_all_ages %>% 
  filter(major_category == "Communications & Journalism" | 
           major_category == "Law & Public Policy" | 
           major_category == "Engineering" |
           major_category == "Computers & Mathematics") %>% 
  ggplot(aes(x = unemployment_rate, y = median)) +
  geom_point(aes(color = major), show.legend = FALSE) +
  facet_grid(rows = vars(major_category), scales = "free_y")
```

Second, there is `facet_wrap()`, which splits the graph into a rectangular grid based on the number of rows or columns that are desired. Unlike  `facet_grid()`, `facet_wrap()` can only use a single variable to split the dataset into a rectangular grid. Figure \@ref(fig:facet-wrap-ex) shows the effect of using `facet_wrap()` with the `major_category` variable. To improve readability, `ncols` will be set to three in this example. More help can be found by typing `?facet_wrap` into the console. 

```{r facet-wrap-ex, eval = TRUE}
college_all_ages %>% 
  ggplot(aes(x = unemployment_rate, y = median)) +
  geom_point(aes(color = major), show.legend = FALSE) +
  facet_wrap(facets = vars(major_category), ncol = 3, scales = "fixed")
```

Additional reading and information about facets can be found [here](https://r4ds.had.co.nz/data-visualisation.html#facets)

## Stats

The previous examples using geometric objects have demonstrated how the data can be used to visualize the data itself. An extension of this is to create a layer based on a statistical calculation. Stats objects do just that. 

The format of geometric objects is `geom_[type]`. The format of stats objects is `stat_[type]`. 

Continuing with the previous example of the unemployment rates and median income of all college major categories, we can use a stats object to overlay the mean onto the scatterplot.

```{r stats-obj-ex, eval = TRUE}
college_all_ages %>% 
  ggplot(aes(x = unemployment_rate, y = median)) +
  geom_point(aes(color = major_category), show.legend = FALSE) + 
  stat_summary(geom = "hline", y.fun = "mean", 
               yintercept = mean(college_all_ages$median), color = "black")
```

Additional reading and information about stats objects can be found [here](https://r4ds.had.co.nz/data-visualisation.html#statistical-transformations)

## Scales

The scale layers allow for the graph scales to be adjusted to specifications. The x and y scales can be set individually through layers such as `scale_x_continuous()`, `scale_y_continuous()`, `scale_x_discrete()`, or `scale_y_discrete()`. The choice of continuous or discrete is dependent on the data type as to which one should be called.

In continuing the example using the college major data, `scale_y_continuous()` and `scale_x_continuous()` can be used to scale the data. Note that continuous was chosen because both the median and the unemployment data are continuous variables.

```{r scale-ex, eval = TRUE}
college_all_ages %>% 
  ggplot(aes(x = unemployment_rate, y = median)) +
  geom_point(aes(color = major_category), show.legend = FALSE) +
  scale_x_continuous(name = "Unemployment rate", limits = c(0, 0.2)) + 
  scale_y_continuous(name = "Median income")
```

## Other geometric objects

While the above example is useful for demonstrating how graphs and figures can be creating using ggplot2, a major limitations is often that creating good graphs is limited by the user's understanding of what graphs are possible with ggplot2. There are dozens of different geometric objects available through ggplot2. Since descriptions of all of them are beyond the scope of this vignette, Hadley Wickham provides a general description of geometric objects [here](https://r4ds.had.co.nz/data-visualisation.html#geometric-objects) and there is a a "Data Visualization Cheat Sheet" available [here](https://www.rstudio.com/resources/cheatsheets/#ggplot2). The Cheat Sheet provides examples of all of the different geometric objects and their arguments as well as different stats and scales options with their respective arguments.

## Additional aesthetic arguments

The above example demonstrates how to create a simple graph using the `geom_point` and `geom_smooth` geometric objects with the *x*, *y*, and *color* aesthetics. The above example also included `geom_smooth` specific arguments for *method* and *se*. Hadley Wickham provides a comprehensive description of all of the aesthetic arguments [here](https://r4ds.had.co.nz/data-visualisation.html#aesthetic-mappings). 

Additional walkthroughs of visualizing data through ggplot2 can be found [here](https://tidy-ds.wjakethompson.com/days/day1/visualize/), where Jake Thompson has created a compilation of examples, demonstrations, and resources of plotting in ggplot2.

## Themes

ggplot comes with a handful of pre-built themes that allow the user to change the appearance of the graph to be in accordance with the theme's settings. The theme can be set by calling the theme's function as an additional layer of the graph. Figure \@ref(fig:theme-ex) sets the previous graph to use the `theme_bw()` theme. 

```{r theme-ex, eval = TRUE}
college_all_ages %>% 
  ggplot(aes(x = unemployment_rate, y = median)) +
  geom_point(aes(color = major_category), show.legend = FALSE) +
  scale_x_continuous(name = "Unemployment rate", limits = c(0, 0.2)) + 
  scale_y_continuous(name = "Median income") +
  theme_bw()
```

# Using theme_atlas()

The first of ratlas' theme packages is `theme_atlas()`, which specifies font characteristics, axes and labels, and background display settings. The `theme_atlas()` theme can be enacted as shown in Figure \@ref(fig:theme-atlas-ex).

```{r theme-atlas-ex, eval = TRUE}
college_all_ages %>% 
  ggplot(aes(x = unemployment_rate, y = median)) +
  geom_point(aes(color = major_category), show.legend = FALSE) +
  scale_x_continuous(name = "Unemployment rate", limits = c(0, 0.2)) + 
  scale_y_continuous(name = "Median income") +
  theme_atlas()
```

# Using theme_atlas_ms()

The second of ratlas' theme packages is `theme_atlas_ms()`, which is similar to `theme_atlas()` but specifies a different font. The `theme_atlas_ms()` theme can be enacted as shown in Figure \@ref(fig:theme-atlas-ms-ex).

```{r theme-atlas-ms-ex, eval = TRUE}
college_all_ages %>% 
  ggplot(aes(x = unemployment_rate, y = median)) +
  geom_point(aes(color = major_category), show.legend = FALSE) +
  scale_x_continuous(name = "Unemployment rate", limits = c(0, 0.2)) + 
  scale_y_continuous(name = "Median income") +
  theme_atlas_ms()
```

# Using the OkabeIto palette

Palettes are assortments of colors that are pre-prepared for use in a graph. The colorblindr package in R is specially developed to provide visualizations that can accomodate those with colorblindness. The OkabeIto palette is one such palette in colorblindr. 

To use the OkabeIto pallete, simply add `scale_fill_OkabeIto(palette = palette_OkabeIto)` as another layer to the graph. This will convert the colors of the graph to those of the color palette. An example is provided in Figure \@ref(fig:okabeito-example).

```{r okabeito-example}
college_all_ages %>% 
  ggplot(aes(x = unemployment_rate, y = median)) +
  geom_point(aes(color = major_category), show.legend = FALSE) +
  scale_x_continuous(name = "Unemployment rate", limits = c(0, 0.2)) + 
  scale_y_continuous(name = "Median income") +
  theme_atlas_ms() + 
  scale_fill_OkabeIto(palette = palette_OkabeIto)
```

There is a second OkabeIto pallete called `palette_OkabeIto_black`, and it can be called with `scale_fill_OkabeIto(palette = palette_OkabeIto_black`.

# Using the atlas palette

Similarly, the atlas palette can be used. The major difference is that the `scale_fill_OkabeIto()` layer is specific to the colorblindr package. For general purposes, a layer calling `scale_fill_continuous()` or `scale_fill_discrete()` can be used depending on whether you data is contintuous or discrete, respectively. An example is provided in Figure \@ref(fig:atlas-palette-example)

```{r atlas-palette-example}
college_all_ages %>% 
  ggplot(aes(x = unemployment_rate, y = median)) +
  geom_point(aes(color = major_category), show.legend = FALSE) +
  scale_x_continuous(name = "Unemployment rate", limits = c(0, 0.2)) + 
  scale_y_continuous(name = "Median income") +
  theme_atlas_ms() + 
  scale_fill_continuous(palette = palette_atlas)
```

Just like the OkabeIto palette, there is a second palette choice for the atlas palettes. Using the same naming format as the OkabeIto palettes, the second atlas palette is titled `palette_atlas_black`, and it can be called using `scale_fill_continuous(palette = palette_atlas_black` or `scale_fill_discrete(palette = palette_atlas_black)`.

# Using theme setting functions

