---
title: "Report Writing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{report-writing-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ratlas)
library(here)
library(tidyverse)
```

## Creating a New Project

When beginning a new report, the first step is to create a new project
in RStudio. This can be done by clicking: 
  
<center> **File > New Project > New Directory > New Project** </center>

From there, specify the project name and the directory where you would like to 
save the report project. To lessen confusion later, it is usually worthwhile to 
save each report in its own directory so that you can quickly and easily find 
the report at a later date.

## Writing the Setup Chunk

## Structuring the Project Directories

## Create and Cross Reference Tables

## Create and Cross Reference Figures

## Using Inline R Code 

## Combining Words

## Turning Numbers Into Words

